Temporal Dynamics of Normalization Reweighting

### 2023-02-10 manuscript.Rmd
Reformatted for journal submmission

---
title: 'Temporal Dynamics of Normalization Reweighting'
author: | 
  | Daniel H. Baker, Daniela Marinova, Richard Aveyard, Lydia J. Hargreaves, Alice Renton,  
  | Ruby Castellani, Phoebe Hall, Miriam Harmens, Georgia Holroyd, Beth Nicholson, 
  | Emily L. Williams, Hannah M. Hobson & Alex R. Wade
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: yes
  word_document: default
  pdf_document:
    toc: no
  html_document: default
bibliography: references.bib
csl: elife.csl
---

@@

# reasonably compact code to check which packages are installed, install the missing ones, and activate all
packagelist <- c('knitr','remotes','tictoc','R.matlab','bookdown','rmarkdown','grImport','tiff','pals','ez','gtools','pracma') # list of CRAN packages
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
if (!'osfr' %in% installed.packages()[,1]){remotes::install_github("centerforopenscience/osfr")}
if (!'FourierStats' %in% installed.packages()[,1]){remotes::install_github("bakerdh/FourierStats")}
packagelist <- c(packagelist,'osfr','FourierStats')
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

For decades, neural suppression in early visual cortex has been thought to be fixed. But recent work has challenged this assumption by showing that suppression can be \emph{reweighted} based on recent history; when pairs of stimuli are repeatedly presented together, suppression between them strengthens. Here we investigate the temporal dynamics of this process using a steady-state VEP paradigm that provides a time-resolved, direct index of suppression between pairs of stimuli flickering at different frequencies (5 and 7Hz). Our initial analysis of an existing EEG dataset (N=100) indicated that suppression increases substantially during the first 2-5 seconds of stimulus presentation (with some variation across stimulation frequency). We then collected new EEG data (N=100) replicating this finding for both monocular and dichoptic mask arrangements in a preregistered study designed to measure reweighting. A third experiment (N=20) used source localized MEG, and found that these effects are apparent in primary visual cortex (V1), consistent with results from neurophysiological work. Because long-standing theories propose inhibition/excitation differences in autism, we also compared reweighting between individuals with high vs low autistic traits, and with and without an autism diagnosis, across our 3 data sets (total N=220). We find no compelling differences in reweighting that are associated with autism. Our results support the normalization reweighting model, and indicate that for prolonged stimulation, increases in suppression occur on the order of 2-5 seconds after stimulus onset.

Suppressive interactions between neurons are ubiquitous in the nervous system, and normalization (or gain control) processes are considered a canonical neuronal computation [@Carandini2011]. Yet for decades the strength of suppression was treated as fixed, largely due to the observation that adapting to one stimulus does not decrease its suppressive potency [@Foley1997; @Freeman2002]. This orthodoxy was recently challenged by a series of innovative studies showing that normalization can be 'reweighted' by recent history [@Westrick2016; @Aschner2018; @Yiltiz2020]. Specifically, when pairs of stimuli are repeatedly presented together, their neural representations come to suppress each other more strongly. This suggests that, far from being fixed, normalization is a dynamic process that is continuously updated by the sensory environment. Here, our objectives were to determine if the timecourse of these changes can be measured non-invasively from the human brain, assess if they occur similarly in distinct suppressive pathways, and determine whether they differ across the population as a function of autistic traits.

Plastic changes within the visual system occur over multiple timescales [see @Webster2015 for a recent review]. In the retina, adaptation to complete darkness takes around 30 minutes [@Lamb2006], and has established physiological substrates (i.e. the regeneration rate of chemicals in different photoreceptor classes). Cortical forms of adaptation to cues such as stimulus contrast [@Blakemore1969], orientation [@Gibson1937] and motion [@Mather2008] can be observed within a few seconds, but also build up over durations on the order of several minutes [@Greenlee1991]. A recent form of binocular plasticity [@Lunghi2011], in which one eye is patched and subsequently contributes more to binocular vision, requires more than an hour. Other types of adaptation have been identified with changes occurring over longer time periods, such as over several hours [@Kwon2009] or days [@Haak2014]. Finally adaptation to specific combinations of orientation and colour [the @McCollough1965 effect] can persist for several weeks [@Jones1975], even with a brief induction period. Previous work on normalization reweighting has involved adapting sequences of around 40 -- 60s [@Aschner2018; @Yiltiz2020], though it is possible that reweighting might occur faster than this, consistent with other types of contrast adaptation.

Normalization itself is not a single process. Multiple suppressive pathways have been identified in the visual system, including between stimuli differing in orientation [@Heeger1992; @Foley1994], eye-of-origin [@Legge1979; @Sengpiel1994] and spatial position [@Petrov2005; @Cannon1991]. At present there is evidence of normalization reweighting between stimuli with orthogonal orientations [@Aschner2018], and adjacent spatial positions [@Yiltiz2020]. We anticipated that interocular suppression should also be subject to reweighting, but that there might be differences in the dynamics across suppressive pathways. This is plausible, given that suppression within and between the eyes has different spatiotemporal tuning [@Meese2009], and dichoptic masking can be reduced by adapting to the mask [@Sengpiel2005; @Li2005; @Baker2007], unlike within-eye masking. Dynamic fluctuations in interocular suppression are a key feature of binocular rivalry [@Wilson2003; @Alais2010], for which autism-related differences have also been reported [@Robertson2013]. By comparing monocular and dichoptic suppression, we are also able to isolate any contribution of early pre-cortical factors. This is because interocular suppression impacts in primary visual cortex, and therefore bypasses earlier retinal and subcortical stages of processing that may contribute to monocular suppression [@Li2005].

Atypical sensory experience is widely reported by individuals on the autism spectrum [@MacLennan2022; @Simmons2009], but the causal mechanisms remain unclear. Typical issues include hypersensitivity to intense stimuli such as loud sounds, bright lights and strong odours or flavours. Yet fundamental measures of sensitivity such as visual acuity [@Tavassoli2011], contrast sensitivity [@Koh2010], and audiometric performance [@Rosenhall1999] are not consistently different from neurotypical controls. Theoretical accounts of sensory differences in autism have long proposed that the balance of inhibition and excitation may be disrupted [@Rubenstein2003; @Rosenberg2015], and there are isolated results that appear consistent with this. For example neural measures of surround suppression are weaker in autism [@Schallmo2020], binocular rivalry is slower [@Robertson2013] and orientation discrimination performance is better [@Dickinson2014]. Conversely, other studies have failed to find group differences using paradigms that should be equally dependent on inhibition, such as orientation repulsion [@VandeCruys2018], repetition suppression [@Utzerath2018] and psychophysical measures of surround suppression [@Sandhu2020]. The evidence for atypical inhibition in autism is therefore currently inconclusive.

Our recent work has identified a potential autism-related difference in gain control using steady-state EEG methods. @Vilidaite2018 measured contrast response functions for flickering stimuli in a large group of neurotypical adults, who were grouped according to their autism quotient [AQ, see @Baron-Cohen2001] score. The responses were comparable between high and low AQ groups at the flicker frequency of the stimulus. However at the second harmonic frequency (twice the flicker frequency), the high AQ group showed weaker responses than the low AQ group. This effect was replicated in an independent sample of 12 adults with an autism diagnosis (compared with neurotypical controls). Because second harmonic responses are caused by nonlinear interactions in the visual system (they are absent in a linear system), and divisive suppression is a nonlinearity, this potentially implicates differences in suppression in autism. Furthermore, there appears to be a developmental trajectory, as autistic children tested as part of the same study [@Vilidaite2018] showed weaker responses at both the first and second harmonic frequencies.

In this paper we perform a time-course analysis of a previously published EEG data set, and report two novel pre-registered experiments using EEG and MEG. Our data show that suppression increases substantially during the first 2-5 seconds following stimulus onset, for both monocular and dichoptic masks. Source localisation of MEG data indicate that the reweighting is present as early as primary visual cortex (V1). We also hypothesised that normalization reweighting might differ as a function of autistic traits. The relative novelty of the reweighting framework could explain why any differences have not previously been detected, and why the literature on inhibition in autism is relatively inconclusive. However we did not find convincing support for this hypothesis.

@@

Taking the ratio of the two timecourses to calculate a masking index reveals that at 7Hz masking increases steeply during the first two seconds of stimulus presentation, and then plateaus for several seconds (blue trace in Figure \@ref(fig:Pilotdata)d). A similar pattern is observed at 5Hz (red trace in Figure \@ref(fig:Pilotdata)d), as well as at the second harmonics, with some variability in the timecourse across frequencies; for example at 5Hz suppression peaks at around 4 seconds. The black trace shows the average masking ratio across all four frequencies, which rises steeply for just over two seconds and then stays approximately constant until stimulus offset. We conducted cluster-corrected t-tests between ratios separated by `r timeoffset`ms, testing for an increase in suppression ratio across time (i.e. a one-sided test). Points at y = 1.8 in Figure \@ref(fig:Pilotdata)d indicate time points where the ratio is significantly increasing, and occur mostly during the first 2 seconds of stimulus presentation.

@@

@@

We compared the timecourse of suppression between groups using a nonparametric cluster correction approach [@Maris2007] to control the type I error rate. Significant clusters are indicated at y = 0.8 in panels d-h of Figure \@ref(fig:AQanalysis). Despite some occasionally significant clusters, there is no clear or consistent difference between groups across our three data sets. In particular, none of the significant clusters occur during the first few seconds of stimulus onset, when reweighting takes place. We also compared suppression ratios calculated on Fourier components for the full trial, and found no significant effects of autism on suppression strength. We therefore conclude that autism/AQ score is not associated with normalization reweighting, or the strength of suppression more generally.

Participants in Experiments 2 & 3 also completed the sensory perception quotient (SPQ) scale. SPQ scores showed significant negative correlation with AQ for both data sets (EEG data, *r* = `r round(corEEG$estimate,digits=2)`, *p* < `r ceiling(1000*corEEG$p.value)/1000`; MEG data, *r* = `r round(corMEG$estimate,digits=2)`, *p* = `r round(corMEG$p.value,digits=3)`) with effect sizes comparable to those reported previously [@Tavassoli2014]. We also conducted an exploratory analysis of the EEG data from Experiment 2, splitting participants by SPQ instead of AQ. However this analysis did not reveal any convincing differences in normalization reweighting either. Our preregistration also proposed to replicate our earlier finding of a reduced second harmonic response in participants with autism/high AQ scores. However the changes to the experimental design greatly reduced the second harmonic response in both experiments, such that it could not be observed reliably (see Figures \@ref(fig:EEGdata)b and \@ref(fig:MEGdata)b). We were therefore not confident in conducting this analysis. We suspect that the increase in spatial frequency from 0.5 c/deg in the @Vilidaite2018 study to 2 c/deg here is most likely responsible for the dramatically reduced second harmonic response.

The idea that the dynamic balance of inhibition and excitation might be different in autism [@Rubenstein2003; @Rosenberg2015] has compelling face validity. For example individuals with autism often report difficulties with changes in their sensory environment, which might be due to gain control processes failing to adapt appropriately. Indeed, there is experimental evidence of reduced adaptation across various domains [@Pellicano2007; @Turi2015], which is predicted by some autism models [@Pellicano2012]. However this appears not to extend to changes in normalization reweighting, despite the link between reweighting and adaptation [@Westrick2016].

# Methods

@@

## Procedures

@@

## Data analysis

@@

## Preregistration, data and code availability

@@

>>>>

---
title: 'Temporal Dynamics of Normalization Reweighting'
author: | 
  | Daniel H. Baker, Daniela Marinova, Richard Aveyard, Lydia J. Hargreaves, Alice Renton,  
  | Ruby Castellani, Phoebe Hall, Miriam Harmens, Georgia Holroyd, Beth Nicholson, 
  | Emily L. Williams, Hannah M. Hobson & Alex R. Wade
date: "Department of Psychology, University of York, UK"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: yes
  word_document: default
  pdf_document:
    toc: no
  html_document: default
bibliography: references.bib
csl: elife.csl
---

@@

# reasonably compact code to check which packages are installed, install the missing ones, and activate all
packagelist <- c('knitr','remotes','tictoc','R.matlab','bookdown','rmarkdown','grImport','tiff','pals','ez','gtools','pracma','osfr') # list of CRAN packages
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
if (!'FourierStats' %in% installed.packages()[,1]){remotes::install_github("bakerdh/FourierStats")}
packagelist <- c(packagelist,'FourierStats')
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

@@

@@

# AQ analysis and figure plotting

@@

For decades, neural suppression in early visual cortex has been thought to be fixed. But recent work has challenged this assumption by showing that suppression can be \emph{reweighted} based on recent history; when pairs of stimuli are repeatedly presented together, suppression between them strengthens. Here we investigate the temporal dynamics of this process using a steady-state visual evoked potential (SSVEP) paradigm that provides a time-resolved, direct index of suppression between pairs of stimuli flickering at different frequencies (5 and 7Hz). Our initial analysis of an existing EEG dataset (N=100) indicated that suppression increases substantially during the first 2-5 seconds of stimulus presentation (with some variation across stimulation frequency). We then collected new EEG data (N=100) replicating this finding for both monocular and dichoptic mask arrangements in a preregistered study designed to measure reweighting. A third experiment (N=20) used source localized MEG, and found that these effects are apparent in primary visual cortex (V1), consistent with results from neurophysiological work. Because long-standing theories propose inhibition/excitation differences in autism, we also compared reweighting between individuals with high vs low autistic traits, and with and without an autism diagnosis, across our 3 data sets (total N=220). We find no compelling differences in reweighting that are associated with autism. Our results support the normalization reweighting model, and indicate that for prolonged stimulation, increases in suppression occur on the order of 2-5 seconds after stimulus onset.

# Significance statement

We investigated the timecourse of a novel form of neural plasticity, normalization reweighting, using steady-state EEG in humans. We find increases in suppression during the first 2-5 seconds of stimulation, with similar dynamics for monocular and dichoptic presentation. The effects are present in primary visual cortex, and were not mediated by autistic traits. These findings demonstrate that reweighting builds up rapidly after stimulus onset, consistent with a process that adjusts the sensory response to the local environment. Normalization reweighting also has potential as a biomarker of neural function both in humans and in animal models of disease.

Suppressive interactions between neurons are ubiquitous in the nervous system, with normalization considered a canonical neuronal computation [@Carandini2011]. Yet for decades the strength of suppression was treated as fixed, due to the observation that adapting to one stimulus does not decrease its suppressive potency [@Foley1997; @Freeman2002]. This orthodoxy has been challenged by a series of innovative studies showing that normalization can be 'reweighted' by recent history [@Westrick2016; @Aschner2018; @Yiltiz2020]. When pairs of stimuli are repeatedly presented together, their neural representations come to suppress each other more strongly. Far from being fixed, normalization is therefore a dynamic process that is continuously updated by the sensory environment. Our objectives were to measure the timecourse of these changes non-invasively in the human brain, compare them across distinct suppressive pathways, and determine whether they differ as a function of autistic traits.

Plastic changes within the visual system occur over multiple timescales [see @Webster2015 for a recent review]. Cortical forms of adaptation to cues such as stimulus contrast [@Blakemore1969], orientation [@Gibson1937] and motion [@Mather2008] can be observed within a few seconds, but also build up over durations on the order of minutes [@Greenlee1991]. Other types of adaptation have been identified where changes occur over longer time periods, such as several hours [@Kwon2009] or days [@Haak2014]. Previous normalization reweighting studies involved adapting sequences of around 40 -- 60s [@Aschner2018; @Yiltiz2020], but in principle reweighting might occur faster than this, consistent with other types of contrast adaptation.

Multiple suppressive pathways have been identified in the visual system, including between stimuli differing in orientation [@Heeger1992; @Foley1994], eye-of-origin [@Legge1979; @Sengpiel1994] and spatial position [@Petrov2005; @Cannon1991]. At present there is evidence of normalization reweighting between stimuli with orthogonal orientations [@Aschner2018], and adjacent spatial positions [@Yiltiz2020]. We anticipated that interocular suppression should also be subject to reweighting, but that there might be differences in the dynamics across suppressive pathways [e.g. @Meese2009; @Sengpiel2005; @Li2005]. Comparing monocular and dichoptic suppression permits any contribution of early pre-cortical factors to be isolated. This is because interocular suppression impacts in primary visual cortex, and bypasses any retinal and subcortical stages of processing that contribute to monocular suppression [@Li2005].

Atypical sensory experience, including hypersensitivity to loud sounds, bright lights and strong odours or flavours, is widely reported by individuals on the autism spectrum [@MacLennan2022; @Simmons2009], but the causal mechanisms remain unclear. Fundamental measures of sensitivity including visual acuity [@Tavassoli2011], contrast sensitivity [@Koh2010], and audiometric performance [@Rosenhall1999] are not consistently different from neurotypical controls. Theoretical accounts of sensory differences in autism have proposed that the balance of inhibition and excitation may be disrupted [@Rubenstein2003; @Rosenberg2015], yet the evidence is currently inconclusive [e.g. @Schallmo2020; @VandeCruys2018; @Sandhu2020]. Our recent work identified an autism-related difference using steady-state EEG [@Vilidaite2018], in which nonlinear (second harmonic) responses were weaker, implicating atypical suppression in autism.

Here we perform a time-course analysis of a previously published data set, and report two novel pre-registered experiments using EEG and MEG. Our data show that suppression increases substantially during the first 2-5 seconds following stimulus onset, for both monocular and dichoptic masks. Source localisation of MEG data indicate that the reweighting is present as early as primary visual cortex (V1). We also hypothesised that normalization reweighting might differ as a function of autistic traits, but did not find convincing support for this hypothesis.

# Materials and Methods

@@

## Experimental design and statistical analysis

@@

@@

## Preregistration, data and code accessibility

@@

Taking the ratio of the two timecourses to calculate a masking index reveals that at 7Hz masking increases steeply during the first two seconds of stimulus presentation, and then plateaus for several seconds (blue trace in Figure \@ref(fig:Pilotdata)d). A similar pattern is observed at 5Hz (red trace in Figure \@ref(fig:Pilotdata)d), as well as at the second harmonics, with some variability in the timecourse across frequencies; for example, at 5Hz suppression peaks at around 4 seconds. The black trace shows the average masking ratio across all four frequencies, which rises steeply for just over two seconds and then stays approximately constant until stimulus offset. We conducted cluster-corrected t-tests between ratios separated by `r timeoffset`ms, testing for an increase in suppression ratio across time (i.e. a one-sided test). Points at y = 1.8 in Figure \@ref(fig:Pilotdata)d indicate time points where the ratio is significantly increasing, and occur mostly during the first 2 seconds of stimulus presentation.

We compared the timecourse of suppression between groups using a nonparametric cluster correction approach [@Maris2007] to control the type I error rate. Significant clusters are indicated at y = 0.8 in panels d-h of Figure \@ref(fig:AQanalysis). Despite some occasionally significant clusters, there is no clear or consistent difference between groups across our three data sets. In particular, none of the significant clusters occur during the first few seconds of stimulus onset, when reweighting takes place. We also compared suppression ratios calculated on Fourier components for the full trial, and found no significant effects of autism on suppression strength. For Experiment 1 we assessed the first and second harmonics separately, but also found no AQ-related differences. We therefore conclude that autism/AQ score is not associated with normalization reweighting, or the strength of suppression more generally.

Autism is composed of a set of heterogenous symptoms and characteristics, and normalization reweighting may have a more specific relationship to some aspects of autism, rather than autism per se. For this reason, we also examined relationships with the sensory perception quotient (SPQ) to examine whether sensory experiences specifically were related to normalization reweighting. SPQ scores showed significant negative correlation with AQ for the data sets from Experiments 2 and 3 (EEG data, *r* = `r round(corEEG$estimate,digits=2)`, *p* < `r ceiling(1000*corEEG$p.value)/1000`; MEG data, *r* = `r round(corMEG$estimate,digits=2)`, *p* = `r round(corMEG$p.value,digits=3)`) with effect sizes comparable to those reported previously [@Tavassoli2014]. We also conducted an exploratory analysis of the EEG data from Experiment 2, splitting participants by SPQ instead of AQ. However this analysis did not reveal any convincing differences in normalization reweighting either. Our preregistration also proposed to replicate our earlier finding of a reduced second harmonic response in participants with autism/high AQ scores. However the changes to the experimental design greatly reduced the second harmonic response in both experiments, such that it could not be observed reliably (see Figures \@ref(fig:EEGdata)b and \@ref(fig:MEGdata)b). We were therefore not confident in conducting this analysis. We suspect that the increase in spatial frequency from 0.5 c/deg in the @Vilidaite2018 study to 2 c/deg here is most likely responsible for the dramatically reduced second harmonic response.

The idea that the dynamic balance of inhibition and excitation might be different in autism [@Rubenstein2003; @Rosenberg2015] has compelling face validity. For example, individuals with autism often report difficulties with changes in their sensory environment, which might be due to gain control processes failing to adapt appropriately. Indeed, there is experimental evidence of reduced adaptation across various domains [@Pellicano2007; @Turi2015], which is predicted by some autism models [@Pellicano2012]. However this appears not to extend to changes in normalization reweighting, despite the link between reweighting and adaptation [@Westrick2016].
